---
title: "sicss-howard-mathematica-R-2022 challenge problem solutions"
author: "Evan Muzzall"
date: '2022-06-16'
output: html_document
---

# Part 1 - Challenge - dataframes

1. Create a dataframe that contains 6 rows and 3 columns by following the instructions above.

```{r}
school <- c("Stanford", "Stanford", "Stanford", "UC Berkeley", "UC Berkeley", "UC Berkeley")
year <- c(1990, 2000, 2020, 1990, 2000, 2020)
pop <- c(10000, 14000, 20000, 20000, 30000, 45000)
income <- year * pop

df <- data.frame(school, year, pop, income)
df
```

2. Advanced: what is the difference between a data frame and a tidyverse tibble? 

https://tibble.tidyverse.org/

# Part 1 - Challenge - subsetting

1. Create a subset that contains data from DC or stops less than or equal to 7000 and just columns "city" and "traffic_stops"

```{r}
traffic_stops <- c(8814, 9915, 9829, 10161, 6810, 8991)
city <- as.factor(c("SF", "DC", "DC", "DC", "SF", "SF"))
year <- as.integer(c(2000, 2000, 2001, 2002, 2001, 2002))
police <- data.frame(city, traffic_stops, year)
police
class(police)

subset3 <- police[police$city == "DC" | police$traffic_stops <= 7000, c("city", "traffic_stops")]
subset3

# 2. Advanced: use the `filter()` and `select()` functions from the `dplyr` R package to do the same thing. 
library(dplyr)
subset3_dplyr <- police %>%
  filter(city == "DC" | traffic_stops <= 7000) %>%
  select(city, traffic_stops)
subset3_dplyr
```

# Part 1 - Challenge - workflows

1. Reproduce the above workflow using the dataset "ca_oakland_2020_04_01 2.csv".

2. Ask the question: how does the distribution of arrests vary by race? Hint: use a barplot (you might have to generate frequencies first to make the barplot!).

```{r eval = F}
?barplot
?t.test
```

```{r}
# Import
oak <- read.csv("data/raw/ca_oakland_2020_04_01 2.csv",
               stringsAsFactors = TRUE)
str(oak)

# Wrangle
oak_sub <- oak[ , c("subject_race", "arrest_made")]
str(oak_sub)
head(oak_sub)

# Summarize
table(oak$arrest_made)

arrests <- table(oak$subject_race, 
                 oak$arrest_made)
arrests

# Visualize
barplot(arrests[,2], las = 2)

# Analyze
oak_aov <- aov(oak$arrest_made ~ oak$subject_race)
summary(oak_aov)
TukeyHSD(oak_aov)
```

3. Save the "ca_oakland_2020_04_01 2.csv" dataset in a variable named oak. Use the `save()` function to save it in a file name "oak.csv" in the "data/preprocessed" folder.

```{r}
save(oak, file = "data/preprocessed//oak.RData")
```

4. Advanced: Use ggplot2 to make the barplot. 

What other statistical applications might be more appropriate? 

- Factor Analysis for Mixed Data: http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/115-famd-factor-analysis-of-mixed-data-in-r-essentials/
- Multiple Correspondence Analysis: http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/114-mca-multiple-correspondence-analysis-in-r-essentials/
- Cramer's V correlation for categorical data: https://rdrr.io/cran/DescTools/man/CramerV.html

```{r}
library(dplyr)
library(ggplot2)

oak_summary <- oak_sub %>%
  group_by(subject_race) %>%
  summarise(count = n())
oak_summary


ggplot(data = oak_summary, aes(x = subject_race, 
                               y = count, 
                               fill = subject_race)) + 
  geom_bar(stat = "identity", 
           width = 0.5, 
           color = "black") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none") +
  xlab("Driver race") + 
  ylab("Number of stops") + 
  ggtitle("Number of stops by driver race,\n Oakland, California, USA\nMarch 2013 - December 2017") 
```

# Part 2 - Challenge - messy data

1. Why is this data not tidy?

```{r}
tidyr::table4a
# ?table4a
```

The two years should be contained within one column. Their values should be contained in a separate column. 


# Part 2 - Challenge - `pivot_longer()`

1.  Why is this data not tidy? (This exercise comes from [`pivot` function vigenette](https://tidyr.tidyverse.org/articles/pivot.html).) Too long or too wide?

```{r}
tidyr::billboard
```

2.  How can you fix it? Which pivot?

```{r}
billboard_ready <- tidyr::billboard %>%
  tidyr::pivot_longer(
    cols = starts_with("wk"), # Use regular expressions
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE # Drop NAs
)
billboard_ready
```

# Part 2 - Challenge - explicit and implicit missing values

1 This example comes from [*R for Data Science*](https://r4ds.had.co.nz/tidy-data.html). Where is the explicit missing value? 

```{r}
stocks <- tibble(
  year = c(2019, 2019, 2019, 2020, 2020, 2020),
  qtr = c(1, 2, 3, 2, 3, 4),
  return = c(1, 2, 3, NA, 2, 3)
)
stocks
```

Use `pivot_wider()` to widen the `stocks` dataframe into a datafrme named `stocks_wide`. Does `stocks_wide` have implicit missing values?

```{r}
# implicit missing values become explicit
stocks_wider <- stocks %>%
  tidyr::pivot_wider(
    names_from = year,
    values_from = return
  )
stocks_wider
```

# Part 2 - Challenge - `filter()`

1. Use `filter(between())` to subset a dataframe that contains only drivers between the ages of 10 and 13. 

```{r}
minors <- clean %>%
  dplyr::filter(dplyr::between(age, 10, 13))
minors
```

# Part 2 - Challenge - combining `group_by()` and `summarize()`

1. Create a new tibble that contains average age and standard deviation by sex and arrest made. 

```{r}
combined_summary <- clean %>%
  group_by(sex, arrest_made) %>%
  summarize(mean_age = mean(age, na.rm = TRUE),
            sd_age = sd(age, na.rm = TRUE))
combined_summary
```

# Part 2 - Challenge - `summarize(across())`

1. use a combination of `summarize(across())` to compute the means of the last three column in the dataframe below. Use the help files and your intuition to figure it out!

```{r}
wide_data <- read.table(header = TRUE,
                        text = "
 patient sex control meas1 meas2
       1   M     8.9  14.1  11.2
       2   F     6.1  11.6  10.3
       3   F     8.2  13.5  14.1
       4   M    12.5  12.4  11.8
")
wide_data

wide_data %>%
  summarize(across(control:meas2, mean, na.rm = TRUE))
```

# Part 2- Challenge - ggplot2

1. Use the `load()` function to load "oak.RData". 

```{r}
## YOUR CODE HERE
load("data/preprocessed/oak.RData")
```

2. Make boxplots for male and female drivers. 

- Define a custom y-axis with intervals of 0, 20, 40, 60, 80, and 100. 
- Fill the boxes with the "BuPu" RColorBrewer palette. 
- Apply the Stata ggtheme

```{r}
## YOUR CODE HERE
ggplot(clean, 
       aes(x = sex, 
           y = age,
           fill = sex)) + 
  geom_boxplot() + 
  scale_y_continuous(breaks = c(0, 20, 40, 60, 80, 100),
                     limits = c(0, 100)) + 
  scale_fill_brewer(palette = "BuPu") + 
  ggthemes::theme_stata() + 
  theme(legend.position = "none")
```